\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-02-27 docs/papers/sto/so.tex
WIP draft loss function

The loss function is constructed by comparing the simulated field $\hat
f$ with the target field $f$.
\YL{
For a field $f$ in either Lagrangian or Eulerian space%, we have thought
%of 3 forms of MSE-inspired loss:
%%
%\begin{equation}
%\cL_f = \frac1{N_a} \ln
%  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
%       {\sum_{\vq'} f(\vq')^2},
%\end{equation}
%%
%where $N_a$ is the number of time steps it takes to generate the snapshot for
%each iteration
%(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
%their Fourier wavevector $\vk$ in both cases).
%
\begin{equation}
\cL_f = \frac1{N_K N_a} \sum_{ij} \ln \biggl[
\frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk, a_j) - f(\vk, a_j) \bigr|^2}
     {\sum_{\vk' \in K_i} | f(\vk', a_j)|^2} + \epsilon \biggr],
\end{equation}
%
where $N_a$ is the number of snapshots and $N_K$ is the number of
$k$-bins.
$\epsilon$ sets the relative threshold tolerance, below which we
consider $\hat f$ in each $K$ accurate enough and the gradients from
that bin less important.
%and
%%
%\begin{equation}
%\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
%\frac{\sum_{k \in K_i} w(\vk)
%      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
%     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
%\end{equation}
%%
%Then the question is which loss we should use for which $f$, to get
%stable and robust results, e.g., no bump for large or small boxes.
%To further limit our scope in this paper, we should probably only
%consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
%velocity $\vv(\vq)$.
We can use $f \in \{\vs(\vq), \delta(\vx), \vv(\vq), (1+\delta)\vv(\vx),
(1+\delta)\vv\vv(\vx)\}$.
}

For the displacement field, we adopt the mean squared error (MSE) loss
that proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
                   {\sum_{\vq'} s(\vq')^2},
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es) below.
% Say that only s loss is not enough even though in principle it is,
% and cite DrewEtAl x2 too

%\YL{Even though the log should be enough for adaptive optimizer and
%fixed sim config, this may create problem when training on all config at
%the same time. So normalizations inside the log is added above and
%below.}

However, in the Eulerian space, an MSE loss is dominated by the small
scale modes.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
Our loss on the overdensity field $\delta$ sums up its MSE relative to
the target field power in logarithmic bins:
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i \ln
\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |\delta(\vk')|^2},
\end{equation}
%
\begin{equation}
\cL_\delta = \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |\delta(\vk')|^2},
\end{equation}
%
where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
factor that suppresses the high $k$ contribution in Eulerian fields such
as the density field (with $w=1$ for the Lagrangian displacement field).
This is because a PM solver gradually loses small scale information step
by step, and without $w$ setting the priority, the loss function could
encourage the network to fit the small scales at the cost of the
intermediate scales.
Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
Lagrangian space, while the density and momentum fields are in the
Eulerian space:
%
\begin{align}
1 + \delta(\vx) &= \int\!\d^3\vq\,
  \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
(1 + \delta)\vv(\vx) &= \int\!\d^3\vq\,
  \vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
(1 + \delta)\vv\vv(\vx) &= \int\!\d^3\vq\,
  \vv\vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
\end{align}
%
which we compute by scattering particles to the choice of grid for loss.

Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
and isotropic, we can show
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, \ln
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
\end{equation}
%
replacing
%
\begin{equation}
\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
\end{equation}
%
that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
when $\hat\delta$ is identical to $\delta$.
Above we choose the binned version instead for the robustness against
the noisy modewise $|\delta(\vk)|^2$ in the denominator.
The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
loss, while our modification is important to balance among all scales so
that the neural network does not focus too much on either high or low
$k$'s.

We combine Lagrangian and Eulerian fields to form the total loss for a
single snapshot:
%
\begin{equation}
\cJ = \cL_\vs + \cL_\delta.
\end{equation}
%

\YZ{We further sum the loss of all snapshots across a single forward
simulation at different times, i.e.
\begin{equation}
\cJ = \sum_{a_i} \cJ |_{a_i}
\end{equation}
}

>>>>

The loss function is constructed by comparing the simulated field $\hat
f$ with the target field $f$.
In this work, we consider two fields in our loss function, the
Lagrangian particle displacement field $\vs(\vq)$ and the Eulerian
density field $\delta(\vx)$.

For the displacement field, the mean squared error (MSE) loss has been
proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
                   {\sum_{\vq'} s(\vq')^2},
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es).
Even though the logarithm should be enough for adaptive optimizer and
fixed simulation configuration, this may create problem in our training
on all configurations at the same time.
So normalizations inside the logarithm is added above and below.
% Say that only s loss is not enough even though in principle it is,
% and cite DrewEtAl x2 too

However, in the Eulerian space, an MSE loss is dominated by the small
scale modes.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
We introduce a loss function that sums up its MSE relative to the target
field power in logarithmic bins:
%
\begin{equation}
  \cL_f = \frac1{N_K} \sum_{i} \ln \biggl[
  \frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
       {\sum_{\vk' \in K_i} | f(\vk')|^2} + \epsilon \biggr],
\label{loss_power_ln}
\end{equation}
%
where $N_K$ is the number of $k$-bins, and $\epsilon$ sets the relative
threshold tolerance, below which we consider $\hat f$ in each $K$
accurate enough and the gradients from that bin less important.
% Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
% and isotropic, we can show
% %
% \begin{equation}
% \cL_\delta \propto \int \!\d\ln\!k\, \ln
% \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
%   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
% \end{equation}
% %
% replacing
% %
% \begin{equation}
% \cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
% \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
%   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
% \end{equation}
% %
% that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
% when $\hat\delta$ is identical to $\delta$.
% Above we choose the binned version instead for the robustness against
% the noisy modewise $|\delta(\vk)|^2$ in the denominator.
% The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
% loss, while our modification is important to balance among all scales so
% that the neural network does not focus too much on either high or low
% $k$'s.

In this work, we use the loss function in \eqref{loss_power_ln} for
both $\vs(\vq)$ and $\delta(\vx)$.
We combine the two fields to form the total loss for a single snapshot,
and further sum over all snapshots in a single forward simulation
%
\begin{equation}
\cJ = \sum_{a_i} \left[\cL_{\vs(a_i)} + \cL_{\delta(a_i)}\right].
\end{equation}
%
